---
title: "PCA"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
install.packages("factoextra")
library(factoextra)
```

## Principal component analysis

Principal component analysis (PCA) applies a mathematical transformation to a dataset with the intention of explaining as much of the variation in the data as possible using a reduced number of dimensions, or principal components.  The first principal component is a dimension that accounts for a maximal portion of the variation in the dataset, and each successive principal component accounts for the remaining variation while being uncorrelated with (orthogonal to) previous components.  

The goal of PCA is to take variables that are somewhat correlated with each other and transform them into uncorrelated dimensions with the intention of revealing patterns in the data.  

By Nicoguaro - Own work, CC BY 4.0, https://commons.wikimedia.org/w/index.php?curid=46871195


```{r}
wdbc <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data", col_names = FALSE)
features <- c("radius", "texture", "perimeter", "area", "smoothness", "compactness", "concavity",
              "concave_points", "symmetry", "fractal_dimension")
names(wdbc) <- c("id", "diagnosis", paste0(features,"_mean"), paste0(features,"_se"),
                 paste0(features,"_worst"))
```



```{r}
wdbc_pca <- wdbc %>%
      select(-1, -2) %>%
      prcomp(center = TRUE, scale = TRUE)
summary(wdbc_pca)
```
```{r}
fviz_eig(wdbc_pca)
```

```{r}
wdbc_pca$x %>% 
      as_tibble() %>%
      ggplot(aes(x = PC1, y = PC2)) +
      geom_point(pch = 1, alpha = 0.5)
```

```{r}
fviz_pca_ind(wdbc_pca, geom.ind = "point", pointshape = 21, 
             pointsize = 2, 
             fill.ind = wdbc$diagnosis, 
             col.ind = "black", 
             palette = "jco", 
             addEllipses = TRUE,
             label = "var",
             col.var = "black",
             repel = TRUE,
             legend.title = "Diagnosis") +
  ggtitle("2D PCA-plot from 30 feature dataset") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
fviz_pca_var(wdbc_pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

